\documentclass[11pt,a4paper,twocolumn]{article}

% === Packages ===
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{url}
\usepackage[numbers]{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, fit, backgrounds, calc, shapes.geometric}

\graphicspath{{figures/}}

% Listings style for code excerpts
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false
}

% === Front matter ===
\title{Entmoot: Layer-2 Group Communication \\
       for AI Agents \\
       with Merkle-Verified Completeness}

\author{%
  Jerry Fanelli\\
  \texttt{jerry.fanelli@gmail.com}%
}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Recent empirical analysis of the Pilot Protocol agent network shows that 626
autonomous agents can self-organize into a small-world social graph on
pairwise trust alone~\citep{calin2026emergent}. Yet the protocol provides
no primitive for group-scale communication: no broadcast, no shared topic
log, no proof that a subscriber has seen every message in a topic of
interest. We present \emph{Entmoot}, a Layer-2 protocol that closes this gap.
It adds multi-party gossip, topic-based subscription, signed group
membership, and Merkle-verified message completeness on top of Pilot's
existing unicast primitives, without waiting for Pilot's in-progress
``custom networks'' feature. We describe the design, a reference Go
implementation across fifteen packages, and an end-to-end canary test that
shows three sandboxed Pilot nodes converging to byte-identical Merkle roots
in under twelve seconds of wall-clock time over real encrypted tunnels.
Entmoot ships a one-command installer and an Agent-Skills-format
skill document, so the protocol is directly consumable by autonomous
agents. Dissemination follows a Plumtree-style eager/lazy spanning
tree with IHAVE/GRAFT/PRUNE repair, and gossip frames are
multiplexed over persistent \texttt{yamux} sessions per peer;
together these keep cross-mesh hop latency sub-second on a live
three-peer deployment across asymmetric consumer networks.
\end{abstract}

% === 1. Introduction ===
\section{Introduction}
\label{sec:introduction}

Six hundred and twenty-six autonomous AI agents found each other without
human intervention on the Pilot Protocol in early 2026. They registered,
handshook, exchanged trust, and formed a network topology with
heavy-tailed degree distribution, 47$\times$ higher clustering than random,
and a giant component spanning two-thirds of the
population~\citep{calin2026emergent}. Nobody designed this. It emerged
from pairwise primitives: each agent could discover an address, negotiate
a trust relationship, and open an encrypted 1-to-1 tunnel to any trusted
peer. What agents did with those tunnels was their own business.

The protocol that sustains that network is deliberately narrow. Pilot
provides an overlay of virtual addresses, reliable encrypted transport,
NAT traversal~\citep{rfc5128}, and a pairwise trust
store~\citep{pilotprotocol}. It does
not provide a group primitive. An agent wishing to address a dozen peers
at once must either maintain a trust edge with each one (a graph whose
edge count scales quadratically in the group size) or rely on a
centralized relay, reintroducing the single point of failure the overlay
was designed to avoid. The Pilot daemon's own \texttt{broadcast} command
today returns, verbatim: ``broadcast is not available yet; custom
networks are WIP.''

This paper presents \emph{Entmoot}, a Layer-2 protocol that adds group
communication to Pilot without modifying it and without waiting for the
in-progress custom-networks feature. An Entmoot group is a signed
append-only roster of member identities, a content-addressable log of
signed messages, and a gossip protocol that fans messages out over
pairwise Pilot streams to random peers drawn from the roster. Every
member maintains a Merkle tree over the messages it holds in a
deterministic topological order; two members with the same set of
messages compute byte-identical roots, so convergence is checkable in
constant space. Subscribers can declare an MQTT-style topic filter and,
in a later version of the protocol, receive a cryptographic proof that
no matching messages are missing.

Our contributions are as follows.

\begin{enumerate}
    \item A design for Layer-2 group communication on Pilot that relies
    only on primitives Pilot already provides in v1.7.2. Group membership
    is captured in signed rosters above the transport, not in Pilot's
    own (in-progress) network membership mechanism.
    \item A Go reference implementation, \texttt{entmootd}, comprising
    fifteen packages and roughly 7,400 lines of non-test Go code
    (14,000 including tests), including framing, replay protection,
    rate limiting, gossip, Merkle trees, and a Pilot-daemon adapter,
    together with an end-to-end canary test in
    which three peers join a group, exchange three signed messages, and
    converge to byte-identical Merkle roots in approximately twelve
    seconds of wall clock, over real encrypted Pilot tunnels spawned as
    subprocesses in the test.
    \item An agent-facing CLI surface (\texttt{join}, \texttt{publish},
    \texttt{tail}, \texttt{info}, \texttt{query}) with local-IPC
    coordination, a one-command installer (\texttt{install.sh}), and an
    Agent-Skills-format skill document
    (\texttt{skills/entmoot/SKILL.md}) that lets OpenClaw-style agents
    install and operate Entmoot without developer intervention.
\end{enumerate}

The rest of the paper is organized as follows. Section~\ref{sec:background}
summarizes Pilot Protocol and positions related work. Section~\ref{sec:design}
describes Entmoot's data model, wire protocol, gossip strategy, and
security posture. Section~\ref{sec:implementation} discusses the
implementation and integration with Pilot's driver API.
Section~\ref{sec:evaluation} presents the canary results.
Section~\ref{sec:related} surveys related primitives and system work.
Section~\ref{sec:future} discusses deferred items and future work.
Section~\ref{sec:conclusion} concludes.

% === 2. Background ===
\section{Background}
\label{sec:background}

\subsection{Pilot Protocol}

Pilot is an overlay network stack for autonomous agents~\citep{pilotprotocol}.
Each participating node generates an Ed25519 identity~\citep{bernstein2012ed25519,rfc8032}
at registration and receives a 48-bit virtual address of the form
\texttt{N:NNNN.HHHH.LLLL}: a 16-bit network ID and a 32-bit node ID within
that network. Network~0 is the global backbone; all agents are members of
it by default. Pilot reserves additional network IDs for purpose-specific
subgraphs; the \emph{Pilot Networks} feature that populates them is in
early-access release as of this writing. A Pilot Network is a
group-scoped \emph{connectivity} primitive: membership grants pairwise
authorization (``address resolution, connection acceptance, and
datagram delivery''~\citep{pilotprotocol}) between members without
requiring $O(N^2)$ bilateral handshakes. It is not a messaging layer:
Pilot Networks do not define a broadcast primitive, an ordered log,
retention, or completeness proofs --- datagrams between network
members are still point-to-point and stateless. Entmoot and Pilot
Networks are therefore complementary rather than overlapping: Pilot
Networks can eventually replace Entmoot's signed roster as the
membership source of truth (\S\ref{sec:future}), while Entmoot
contributes the broadcast semantics --- fanout, Merkle-verified
completeness, and anti-entropy reconciliation --- that a Network
membership primitive alone does not provide.

Communication is trust-gated. Two nodes must complete a cryptographic
handshake (port~444) mediated by the registry before they can exchange
traffic on any other port~\citep{calin2026emergent}. Once trust is
established, nodes may open a reliable byte stream to each other on any
port. Under the hood, Pilot runs a TCP-equivalent reliable-delivery
protocol with X25519 key exchange~\citep{rfc7748} and AES-256-GCM
authenticated encryption~\citep{nist80038d} per tunnel, the same
primitives that underlie WireGuard~\citep{donenfeld2017wireguard}. The bytes that carry a tunnel move through a
pluggable transport layer: UDP is the default byte-mover and is what
UDP-capable peers use end-to-end, but the fork we build against also
ships a TCP transport that a public-IP daemon can expose alongside UDP
so peers on UDP-hostile networks (restrictive corporate firewalls,
carrier-grade NAT, UDP-blocking ISPs) can still establish tunnels. The
transport choice is invisible above the tunnel layer: the same wire
frames travel over whichever transport the dialing peer's direct-UDP
retries settle on. The registry does not see tunnel payloads; it only
mediates address allocation, hostname resolution, the handshake relay,
and, when present, the per-peer list of transport endpoints.

Crucially, the registry hides the network endpoint of any node not
flagged as ``public.'' A private node's address can only be resolved to
a reachable endpoint by another node with an existing trust edge. A new
participant cannot therefore bootstrap into an otherwise-private group
without either an out-of-band introduction or at least one public
introducer.

We build against the \texttt{jerryfane/pilotprotocol}
fork~\citep{pilotprotocol}, which layers a pluggable transport
abstraction and a cluster of reliability fixes on top of upstream
v1.7.2: NAT-drift-resilient endpoint refresh (a STUN-style
guard against carrier NAT remapping~\citep{rfc8489}), TURN-style
relay routing for peers behind symmetric NAT~\citep{rfc8656},
and a self-dial guard at the transport layer
(\S\ref{sec:security}). The abstraction and fixes are
wire-compatible with unmodified upstream daemons. The companion
evolution paper~\citep{entmootevolution} records the
incident-level narrative of how these were surfaced in live
deployment and shaped the fanout and multiplexing policies
described in~\S\ref{sec:design}.

\subsection{Autonomous agent networks}

Classical multi-agent systems literature~\citep{wooldridge2009,shohamleyton2008,dorri2018}
studies small populations with pre-specified communication protocols,
agent architectures, and hard-coded interaction topologies.
\citet{calin2026emergent} observed that autonomous agents on Pilot, given
only a trust-gated transport, self-organize into structures with
statistical signatures familiar from human social networks: heavy-tailed
degree distribution consistent with preferential
attachment~\citep{barabasi1999}, clustering coefficient $\sim$47$\times$
above random, and small-world
properties~\citep{watts1998small}. Functional clusters (data/analytics,
wellness, career, engineering) emerge without coordination. That study
explicitly flags cross-network community structure as out of reach because
no such structure yet exists in the observed data: all 626 agents were on
network~0. Entmoot instantiates the missing substrate.

\subsection{Primitives we build on}

We reuse ideas from several well-understood distributed-systems
primitives. Topic syntax follows the slash-separated hierarchical form
with \texttt{+} and \texttt{\#} wildcards defined by
MQTT~\citep{mqtt5}. Our Merkle log construction is directly influenced
by Certificate Transparency, including domain-separated leaf and internal
hashes~\citep{laurie2013ct}. Content-addressing by hash echoes
IPFS~\citep{benet2014ipfs}. For comparison to structured peer-to-peer
overlays we cite Kademlia~\citep{maymounkov2002kademlia}, though
Entmoot does not use a DHT.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  every node/.style={draw, rounded corners=1pt, align=center,
                     inner sep=3pt, minimum height=6mm},
  box/.style={minimum width=62mm},
]
\node[box, fill=gray!5]  (cli)   {CLI clients: \texttt{publish},
      \texttt{tail}, \texttt{info}, \texttt{query}};
\node[box, fill=gray!5, below=1mm of cli]  (csock)
     {{\scriptsize control socket
       (\texttt{\textasciitilde/.entmoot/control.sock})}};
\node[box, fill=gray!20, below=1mm of csock]   (ent)
     {\textbf{entmootd join} on port 1004\\
      {\scriptsize wire $\cdot$ roster $\cdot$ store $\cdot$ gossip $\cdot$ merkle $\cdot$ ipc}};
\node[box, fill=gray!5, below=1mm of ent]   (psock)
     {{\scriptsize Pilot IPC socket}};
\node[box, fill=gray!20, below=1mm of psock]  (pilot)
     {\textbf{Pilot daemon}\\
      {\scriptsize identity $\cdot$ trust $\cdot$ tunnels $\cdot$ NAT}};
\node[box, fill=gray!5, below=1mm of pilot] (net)
     {UDP over IP};
\end{tikzpicture}
\caption{System stack and multi-process layout. A single long-running
\texttt{entmootd join} process per host owns the Pilot connection and
the on-disk state; short-lived CLI clients (\texttt{publish},
\texttt{tail}, \texttt{info}, \texttt{query}) dial the local control
socket. Entmoot reaches its peers by opening Pilot streams on
port~1004, with the Pilot daemon providing identity, trust, and
transport encryption.}
\label{fig:architecture}
\end{figure}

% === 3. Design ===
\section{Design}
\label{sec:design}

\subsection{Goals and non-goals}

Figure~\ref{fig:architecture} shows where Entmoot sits in the stack.
Entmoot aims to provide many-to-many communication with selective
synchronization: each member stores only messages matching its topic
filter, yet can prove cryptographically that it has not missed any
in-scope message. The network is fully decentralized; any member can
act as a gossip relay. Entmoot is built on Pilot's primitives as they
exist today, without dependence on custom networks.

Entmoot does not attempt Byzantine consensus on message ordering; it
provides eventual consistency with Merkle-verified completeness, not a
totally ordered log. Group content is author-signed plaintext:
Pilot's transport gives pairwise forward secrecy, but members see each
other's traffic in the clear. We do not design a human-facing interface,
and we do not replace any existing Pilot primitive.

\subsection{Data model}

A \emph{group} is a 32-byte random identifier, a signed membership
roster, and a content-addressed log of messages. A \emph{message} is an
author-signed record containing a topic list, a payload, and up to three
\emph{parent} message identifiers that link it to the messages the
author had seen at composition time. Messages therefore form a DAG, not
a linear log. The cap on parents bounds frame size while preserving
causal ordering; a genesis message has an empty parent list.

An invite bundle is a signed artifact produced by an existing member. It
carries the group identifier, the founder's identity, the current
roster head, a snapshot of the Merkle root, and a list of three to five
recently-online bootstrap peers. It is delivered out-of-band and is
verified by the joiner against the issuer's already-trusted Entmoot
public key.

\subsection{Membership}

The roster is a signed append-only log. Operations are \texttt{add},
\texttt{remove}, and \texttt{policy\_change}. Each entry is signed by
its actor, carries a timestamp, and references the current roster head
in its \texttt{parents} field. The group founder is the only
authorized admin: any entry whose actor is not the founder, or whose
signature does not verify against the founder's public key, is rejected.
This reduces the design to a single-writer append-only log and eliminates
conflict resolution. The upgrade path to multi-admin is documented but
deferred: a policy blob names $n$ admins with a quorum rule such as
$k$-of-$n$, and a deterministic tiebreaker where the lower
\texttt{node\_id} wins on timestamp equality.

\subsection{Wire protocol}

Connections use Pilot's reliable byte streams on a fixed port (we use
1004). Each frame carries a 4-byte big-endian length prefix, a 1-byte
message-type tag, and a JSON body, with a 16\,MiB cap. JSON keeps the
protocol debuggable and extensible; a binary codec is documented as
future work.

The protocol defines thirteen message types: eight for the core
data plane (\texttt{hello}, \texttt{roster\_req},
\texttt{roster\_resp}, \texttt{gossip}, \texttt{fetch\_req},
\texttt{fetch\_resp}, \texttt{merkle\_req}, \texttt{merkle\_resp}),
two for anti-entropy reconciliation (\texttt{range\_req},
\texttt{range\_resp}; \S\ref{sec:antientropy}), and three for
Plumtree tree repair (\texttt{ihave}, \texttt{graft},
\texttt{prune}; \S\ref{sec:gossip}).%
\footnote{The \texttt{gossip} frame carries an optional inline
message body, used when the canonical-encoded \texttt{Message} is at
most 4\,KiB; the receiver hash-verifies the body against the
advertised id before storing. All frames between two peers are
multiplexed over a single per-peer
\texttt{yamux}~\citep{hashicorpyamux} session opened on top of
Pilot's stream; yamux framing is invisible above the transport
adapter (\S\ref{sec:implementation}).}
Every type that carries a timestamp is subject to replay protection
that mirrors Pilot's own handshake protocol: reject messages older than
five minutes or more than thirty seconds in the future, and dedupe by
body SHA-256 in a bounded LRU set.

\subsection{Bootstrap}

A new member joining an existing group exercises a three-tier strategy,
tried in order:

\begin{enumerate}
    \item \textbf{Invite bundle (primary).} The joiner dials each peer
    listed in the invite's \texttt{bootstrap\_peers} on port~1004 until
    one responds. On first success, it issues a \texttt{roster\_req} and
    applies the returned entries.
    \item \textbf{Pilot-trusted peers intersected with the roster
    (secondary).} If every bootstrap peer is offline, the joiner queries
    its local Pilot daemon for its trusted-peers set and intersects with
    roster members. Any member we already Pilot-trust is a legitimate
    candidate.
    \item \textbf{Founder fallback (tertiary).} The founder's node ID
    is always in the roster (it is the genesis entry). As a last
    resort, the joiner dials the founder directly.
\end{enumerate}

Entmoot neither performs DHT-style peer discovery nor registers groups
with Pilot's registry. The out-of-band invite plus Pilot's existing
trust graph are sufficient to bootstrap.

\subsection{Gossip}
\label{sec:gossip}

Dissemination is epidemic in the classical rumour-spreading sense:
push gossip converges in $O(\log N)$ rounds with high
probability~\citep{karp2000rumor,demers1987epidemic}. Entmoot uses
a Plumtree-style~\citep{leitao2007plumtree} spanning tree over a
HyParView~\citep{leitao2007hyparview} peer-sampling substrate.
Every member maintains two peer sets drawn from the roster: an
\emph{eager} set, over which full \texttt{gossip} frames travel,
and a \emph{lazy} set, over which only message identifiers
(\texttt{ihave}) are advertised. Receipt of a duplicate full-body
frame \texttt{prune}s the sender, demoting it to the lazy set so
the redundant edge is pruned from the spanning tree. Receipt of an
\texttt{ihave} for an unknown identifier, with no corresponding
body arriving within a three-second grace interval, triggers a
\texttt{graft} to the advertiser, which replies with the body and
promotes the requester back into its eager set. In steady state
the spanning tree thins to one full-body send per broadcast per
spanning-tree edge.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  lane/.style={draw, rounded corners=1pt, minimum width=13mm,
               minimum height=5mm, inner sep=2pt, fill=gray!15},
  note/.style={draw=none, align=center, font=\scriptsize\itshape},
  msg/.style={-{Latex[length=1.6mm]}, thick},
  graft/.style={-{Latex[length=1.6mm]}, thick, dashed, gray!60!black},
]
\node[lane] (A) at (0, 0)    {$A$};
\node[lane] (B) at (2.1, 0)  {$B$};
\node[lane] (C) at (4.2, 0)  {$C$};

\draw[dashed, gray!60] (A.south) -- ++(0, -5.2);
\draw[dashed, gray!60] (B.south) -- ++(0, -5.2);
\draw[dashed, gray!60] (C.south) -- ++(0, -5.2);

\draw[msg] ($(A.south) + (0, -0.55)$)
        -- node[above, font=\scriptsize] {\texttt{gossip}\,(body)}
           ($(B.south) + (0, -0.55)$);
\draw[msg] ($(A.south) + (0, -1.3)$)
        -- node[above, font=\scriptsize] {\texttt{ihave}\,(id)}
           ($(C.south) + (0, -1.3)$);

\node[note] at (4.2, -2.1)
  {graft timeout (3\,s)};

\draw[graft] ($(C.south) + (0, -3.1)$)
        -- node[above, font=\scriptsize] {\texttt{graft}\,(id)}
           ($(A.south) + (0, -3.1)$);
\draw[msg] ($(A.south) + (0, -3.85)$)
        -- node[above, font=\scriptsize]
           {\texttt{gossip}\,(body); $C$ promoted to eager}
           ($(C.south) + (0, -3.85)$);

\node[note] at (2.1, -4.75)
  {\texttt{prune} fires on duplicate arrival;\\demotes the sender to lazy};

\end{tikzpicture}
\caption{Plumtree spanning-tree repair. $A$ eager-pushes the full
message body to $B$ and a lazy \texttt{ihave} advertisement to
$C$. If the body has not arrived at $C$ within 3~seconds, $C$ sends
\texttt{graft} to $A$, which replies with the body and promotes $C$
back into its eager set. Symmetrically, a duplicate \texttt{gossip}
frame triggers \texttt{prune}, demoting the redundant sender to lazy
so subsequent messages arrive as \texttt{ihave}-only.}
\label{fig:plumtree}
\end{figure}

The first-seen forwarding invariant is applied uniformly across
every acquisition path. A message acquired via eager push, via
retry of an initially-failed fetch, or via the anti-entropy
reconciliation of \S\ref{sec:antientropy} is re-fanouted exactly
once on first sight; the hook lives inside \texttt{fetchFrom}, the
single code path every acquisition funnels through. This matches
Plumtree's canonical rule~\citep{leitao2007plumtree} and the same
invariant applied by libp2p
GossipSub~\citep{vyzovitis2020gossipsub}, and is what lets a
message reach members on the far side of a partial-connectivity
cut via a relay member's re-fanout.

Fanout itself is trust-aware and bounded. Before dispatching a
push or \texttt{ihave}, the gossiper intersects its eager set
with Pilot's per-peer trust table (cached for one second) and
drops peers with no trust edge, which would otherwise time out
at the transport layer. This mirrors libp2p's
\texttt{host.Network().Connectedness()} check. The remaining
dispatches run concurrently under
\texttt{golang.org/x/sync/errgroup} with a per-peer five-second
\texttt{context.WithTimeout}, so one slow peer does not
head-of-line-block the others. A per-peer dial-backoff cache
short-circuits further dials to a peer that has recently failed;
the cache window doubles from one second up to a five-minute
cap, preventing the retry scheduler from burning budget on a
known-dead peer.

Transient \texttt{Transport.Dial} failures are requeued rather
than dropped. A small \texttt{retryLoop} goroutine drains the
pending queue every 500\,ms with
decorrelated-jitter~\citep{brooker2015jitter} backoff, computed
as \texttt{next = min(cap, random(base, prev$\times$3))}. Ten
attempts within a roughly five-minute budget are allowed; slots
that exhaust the budget log the failure and hand off to the
anti-entropy path of \S\ref{sec:antientropy}. Jittered delays
prevent retries from synchronising across the fleet during
correlated blips such as a registry reconnect.

Two wire-level refinements reduce per-hop round trips. The
\texttt{gossip} frame carries the full \texttt{Message} inline
when its canonical encoding is at most 4\,KiB (matching IPFS
Bitswap's default \texttt{WantHaveReplaceSize} threshold), so
the receiver hash-verifies the body and stores directly without
a \texttt{fetch\_req}/\texttt{fetch\_resp} round-trip; larger
messages still pull the body. Separately, all gossip streams
between two peers are multiplexed over a single long-lived
\texttt{yamux}~\citep{hashicorpyamux} session opened on top of
the Pilot tunnel, so opening a new stream is a zero-round-trip
local call; \S\ref{sec:implementation} covers the adapter
implementation.

Push is the fast path. When every gossip recipient is reachable
and willing, fanout~two suffices for a three-member topology to
converge in a single round. The slow path, which handles the
reconnect-missed-gossip case, is \S\ref{sec:antientropy}.
Near-peer sampling informed by subscriber interest overlap is a
documented upgrade, not yet implemented.

\subsection{Anti-entropy reconciliation}
\label{sec:antientropy}

Push gossip, even with the Plumtree spanning-tree repair of
\S\ref{sec:gossip}, is vulnerable to a specific failure mode: a
member that is offline when a message is published, and reconnects
after the push fanout has decayed, never hears about the message
from the push path. A na\"ive fix (have every member periodically
push its entire store to every peer) is bandwidth-ruinous at any
non-trivial group size. Entmoot instead takes the anti-entropy
reconciliation approach introduced by Demers et
al.~\citep{demers1987epidemic} and familiar from Dynamo's
Merkle-tree repair~\citep{decandia2007dynamo}.

Reconciliation is triggered by peer reconnect, not by a periodic timer:
both an inbound \texttt{Accept} and a successful outbound dial invoke
\texttt{maybeReconcile} on the peer, which is rate-limited to one round
per peer per 60 seconds. A round proceeds in four steps routed through
the same retry scheduler the gossip path uses:

\begin{enumerate}
    \item \textbf{Cheap check.} Request the peer's Merkle root via
    \texttt{MerkleReq} and compare to the local root. If they match, the
    two stores are byte-identical under the topological-order rule
    (\S\ref{sec:merkle}), and the round ends in constant space and one
    round-trip. This is the common case after a brief disconnect.
    \item \textbf{Bounded enumeration.} If the roots differ, issue
    \texttt{RangeReq} with a lower timestamp bound set to the local
    store's latest message timestamp. The peer answers with
    \texttt{RangeResp}, a list of message identifiers it holds in that
    window.
    \item \textbf{Targeted fetch.} For each id in \texttt{RangeResp} we
    do not already have, issue an ordinary \texttt{FetchReq}.
    \item \textbf{Signature-verified store.} Each fetched body is
    verified against the author's roster entry before it lands in the
    local store, identical to the push path.
\end{enumerate}

Timestamp-bounding the range query matters: a reconnecting peer is
most likely missing only the tail of the message DAG, so the range
enumeration is cheap. A full-store enumeration remains possible by
choosing a zero bound but is not used by the automatic path. The two
wire types \texttt{MsgRangeReq} (0x09) and \texttt{MsgRangeResp}
(0x0A) are additive: peers that do not speak them reject the frame
as an unknown type and the caller silently falls back to push-only,
preserving backward compatibility with older implementations.
Figure~\ref{fig:antientropy} sketches the message sequence for a
single reconciliation round.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  lane/.style={draw, rounded corners=1pt, minimum width=20mm,
               minimum height=5mm, inner sep=2pt, fill=gray!15},
  note/.style={draw=none, align=left, font=\scriptsize\itshape},
  msg/.style={-{Latex[length=1.6mm]}, thick},
  self/.style={-{Latex[length=1.4mm]}, thick, gray!60!black},
]
% Lane headers
\node[lane] (L) at (0, 0)    {local};
\node[lane] (R) at (4.2, 0)  {remote};

% Lifelines
\draw[dashed, gray!60] (L.south) -- ++(0, -6.8);
\draw[dashed, gray!60] (R.south) -- ++(0, -6.8);

% Step 1: MerkleReq -> MerkleResp
\draw[msg] ($(L.south) + (0, -0.6)$)
        -- node[above, font=\scriptsize] {\texttt{MerkleReq}}
           ($(R.south) + (0, -0.6)$);
\draw[msg] ($(R.south) + (0, -1.35)$)
        -- node[above, font=\scriptsize] {\texttt{MerkleResp}\,(root)}
           ($(L.south) + (0, -1.35)$);

% Branch: roots match → done (short-circuit)
\node[note] at (2.1, -2.0)
  {if roots match: \textbf{done} (constant space)};

% Step 2: RangeReq -> RangeResp (roots differ)
\draw[msg] ($(L.south) + (0, -2.85)$)
        -- node[above, font=\scriptsize]
           {\texttt{RangeReq}\,(since\,$t_\text{local}$)}
           ($(R.south) + (0, -2.85)$);
\draw[msg] ($(R.south) + (0, -3.6)$)
        -- node[above, font=\scriptsize] {\texttt{RangeResp}\,(ids)}
           ($(L.south) + (0, -3.6)$);

% Step 3: diff ids (self-loop on local)
\draw[self, rounded corners=1pt]
        ($(L.south) + (0, -4.25)$)
     -- ++(-0.7, 0) -- ++(0, -0.5) -- ++(0.7, 0);
\node[note, anchor=east] at ($(L.south) + (-0.85, -4.5)$)
  {local:\\diff ids};

% Step 4: FetchReq -> FetchResp (per missing id)
\draw[msg] ($(L.south) + (0, -5.35)$)
        -- node[above, font=\scriptsize]
           {\texttt{FetchReq}\,(id)}
           ($(R.south) + (0, -5.35)$);
\draw[msg] ($(R.south) + (0, -6.1)$)
        -- node[above, font=\scriptsize]
           {\texttt{FetchResp}\,(body)}
           ($(L.south) + (0, -6.1)$);

\node[note] at (2.1, -6.7)
  {repeat per missing id; verify signature};
\end{tikzpicture}
\caption{Anti-entropy reconciliation round between two peers. The cheap
Merkle-root check short-circuits the round in constant space when the
two stores already agree; only when roots differ does the round pay the
cost of enumerating the timestamp-bounded range and fetching the
missing bodies.}
\label{fig:antientropy}
\end{figure}

\subsection{Merkle completeness}
\label{sec:merkle}

Every member maintains a Merkle tree over the set of message identifiers
in its local store, ordered by the topological order defined over the
message DAG. Leaves are hashed as
$\mathrm{sha256}(\mathtt{0x00} \Vert \text{id})$ and internal nodes as
$\mathrm{sha256}(\mathtt{0x01} \Vert \text{left} \Vert \text{right})$.
The domain-separation bytes follow RFC~6962~\citep{laurie2013ct} and
defend against second-preimage attacks. Odd-sized layers promote the
lone node unchanged to the next layer (Figure~\ref{fig:merkle}).

Two members with the same message set under the same ordering rule
compute byte-identical roots. Root comparison is therefore a constant-
space convergence check. Positive-inclusion proofs are implemented;
absence proofs for filtered subscriber views are future work.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  leaf/.style={draw, rectangle, rounded corners=1pt,
               minimum width=9mm, minimum height=5mm, inner sep=1pt},
  internal/.style={leaf, fill=gray!10},
  rootnode/.style={leaf, fill=gray!25}
]
\node[leaf] (l0) at (0, 0)    {$L_0$};
\node[leaf] (l1) at (1.3, 0)  {$L_1$};
\node[leaf] (l2) at (2.6, 0)  {$L_2$};

\node[internal] (n01) at (0.65, 1.2) {$N_{01}$};
\node[internal] (l2p) at (2.6, 1.2)  {$L_2$};

\node[rootnode] (r) at (1.625, 2.4) {\textbf{root}};

\draw (l0) -- (n01);
\draw (l1) -- (n01);
\draw (l2) -- (l2p);
\draw (n01) -- (r);
\draw (l2p) -- (r);

\node[below=1.5mm of l1, draw=none, font=\scriptsize]
  {$L_i = H(\texttt{0x00}\,\|\,m_i.\textsf{id})$};
\node[above=0.5mm of r, draw=none, font=\scriptsize]
  {$H(\texttt{0x01}\,\|\,N_{01}\,\|\,L_2)$};
\end{tikzpicture}
\caption{Merkle tree over three message IDs in topological order. Leaves
are domain-separated by a \texttt{0x00} prefix; internal nodes by
\texttt{0x01}. Odd-layer singletons are promoted unchanged: here $L_2$
has no sibling at level 1 and is lifted to level 2 as-is.}
\label{fig:merkle}
\end{figure}

\subsection{Security posture}
\label{sec:security}

Every state-changing message is signed by its author with an Ed25519
key bound to the author's roster entry. Unsigned or incorrectly-signed
messages are silently dropped. The wire layer enforces a 5-minute-past,
30-second-future replay window plus an 8192-entry LRU of body hashes to
dedupe seen messages. On accept, the server verifies that the remote
peer (as identified by Pilot's trust layer) is a current roster member;
non-members are disconnected before any application logic runs.

Per-peer token-bucket rate limits guard against abuse by a misbehaving
member. The defaults are 100 messages per second with a burst of
200, and 1\,MiB/s with a burst of 4\,MiB. A sustained breach is handled
by hard disconnect; cooldown and soft-penalty schemes are deferred.

Group content is \emph{not} encrypted from other members. Pilot's
transport encryption protects content from outsiders and relays, but
any member sees the plaintext of every message it receives or forwards.
End-to-end group encryption with shared-key rotation on churn is
documented as future work; the framing is arranged so encryption can
be added underneath without breaking protocol compatibility.

Two layered defences guard against self-dial amplification. The
pathology arose in a live deployment on a multi-homed host whose
Docker-bridge and public-IP interfaces both matched Pilot's
same-LAN detection: the daemon established three duplicate
self-tunnels and sustained a 5\,900-packet-per-second retransmit
loop. At the Entmoot layer, the invite minter excludes the
issuer's node id from \texttt{bootstrap\_peers} so no receiver's
Join path ever dials self. At the transport layer, Pilot rejects
any \texttt{DialConnection} whose destination is the local node
id with a typed \texttt{ErrDialToSelf} sentinel, mirroring
\texttt{go-libp2p-swarm}'s canonical guard. Either fix alone
closes the amplification loop; both ship together for defence in
depth.

% === 4. Implementation ===
\section{Implementation}
\label{sec:implementation}

\texttt{entmootd} is implemented in Go (${\geq}1.25.3$, matching
Pilot's toolchain requirement) and compiles against the
\texttt{jerryfane/pilotprotocol} fork of upstream Pilot v1.7.2,
which adds the pluggable transport layer discussed in
\S\ref{sec:background} and the reliability fixes described where
relevant in this section. The module is split into fifteen
packages totalling roughly 7,400 lines of non-test Go code (14,000
including tests). Key packages: \texttt{canonical} (deterministic JSON encoding
used for signing and identifier computation), \texttt{merkle} (tree
construction and positive-inclusion proofs), \texttt{roster} (signed
log with the founder-only admin rule and a dedicated
\texttt{AcceptGenesis} entry point for joiners that lack the founder's
private key), \texttt{store} (in-memory, JSONL, and a SQLite
backend using WAL mode with one database file per group, sharing a
topological-order helper), \texttt{wire} (framing, codec, replay
protection), \texttt{gossip} (push strategy, peer sampling, and the
three-tier bootstrap), \texttt{transport/pilot} (the adapter over
Pilot's Go driver API), and \texttt{ipc} (local control-socket codec,
reusing the framing shape in a separate type namespace so the
security model of peer-facing wire traffic stays split from
same-host IPC). See \texttt{docs/CLI\_DESIGN.md} \S4 for the SQLite
layout and \S5 for the IPC contract.

The binary exposes a five-command agent-facing surface plus two
founder-only subcommands. \texttt{join} consumes an invite, binds the
listen port, opens a local control socket, and enters the accept loop,
blocking until signalled; \texttt{publish} authors and gossips a
message; \texttt{tail} streams live messages with optional SQLite
backfill; \texttt{info} emits a JSON status snapshot; and
\texttt{query} runs indexed historical lookups against SQLite. The
founder-only commands \texttt{group create} and \texttt{invite create}
produce new group identifiers and signed invite bundles respectively.
Default paths follow Pilot's conventions: \texttt{/tmp/pilot.sock} for
Pilot IPC, \texttt{\textasciitilde/.entmoot/} for persistent state,
and \texttt{\textasciitilde/.entmoot/control.sock} for the local
control socket.

The control-socket IPC model keeps multi-process operation coherent.
Exactly one \texttt{join} process per host owns the single Pilot
connection and the SQLite writer for every joined group. Short-lived
CLI invocations (\texttt{publish}, \texttt{tail} in live mode) dial
\texttt{\$\{data\}/control.sock}, send one request, and read one or
more response frames; \texttt{info} and \texttt{query} read SQLite
directly and work whether or not \texttt{join} is running. This layout
enforces ``one Pilot session per identity'' and routes every write
through a single owner, eliminating split-brain replay state and
conflicting roster views. The \texttt{ipc} package defines the wire
framing (4-byte length prefix, 1-byte type, JSON body) and a small
error-code vocabulary that the client translates to process exit
codes; \texttt{docs/CLI\_DESIGN.md} \S5 pins the full contract.

Agent-facing packaging is the final piece. \texttt{install.sh} fetches
prebuilt binaries from GitHub Releases with a source-build fallback
when no release matches the host triple, placing the binary under
\texttt{\textasciitilde/.entmoot/bin}. \texttt{skills/entmoot/SKILL.md}
is a declarative Agent-Skills document that agents read to verify
prerequisites (\texttt{entmootd} and \texttt{pilot-daemon} on
\texttt{\$PATH}, a reachable Pilot socket), install missing
components, and drive the five-command surface.

The Pilot adapter implements a small transport interface that gossip
uses internally:

\begin{lstlisting}[language=Go,basicstyle=\ttfamily\scriptsize]
type Transport interface {
    Dial(ctx context.Context,
         peer NodeID) (net.Conn, error)
    Accept(ctx context.Context) (
         net.Conn, NodeID, error)
    TrustedPeers(ctx context.Context)
         ([]NodeID, error)
    Close() error
}
\end{lstlisting}

The in-memory transport used in unit tests and the Pilot-backed
transport share the same surface. Three integration quirks of the
Pilot driver are worth reporting. First, Pilot wraps the remote
address on a \texttt{net.Conn} in an unexported type, so we resolve
peer node IDs by round-tripping through \texttt{RemoteAddr().String()}
and \texttt{protocol.ParseSocketAddr}. Second, the IPC response for
trusted peers uses the JSON key \texttt{"trusted"}, not \texttt{"peers"},
which is not documented. Third, because the
registry hides private nodes' endpoints from non-trusted lookups, at
least one participant in any bootstrap must be launched with
Pilot's \texttt{-public} flag; otherwise the handshake request
cannot be delivered. Entmoot is otherwise agnostic to which transport
carries a given Pilot frame: when the dialing peer's direct-UDP retries
time out and the callee has advertised a TCP endpoint via the registry,
the Pilot fork falls back to TCP transparently, and the \texttt{net.Conn}
the adapter hands us is indistinguishable from its UDP equivalent.

One operational footnote is worth flagging for readers who will
run \texttt{entmootd} or \texttt{pilot-daemon} on recent macOS.
Ad-hoc-signed Go binaries on macOS~26 (Tahoe) stall at
\texttt{\_dyld\_start} on first interactive launch until the local
\texttt{com.apple.quarantine} extended attribute is cleared. The
fork's \texttt{install.sh} re-signs the installed binaries with
the host's ad-hoc identity and runs \texttt{xattr -cr} on the
binary directory as a post-install step, so the Gatekeeper
first-launch gate passes without user interaction.

A subtlety surfaced by long-running canaries deserves a mention.
Pilot's encrypted tunnels are rekeyed periodically (at Pilot's
\texttt{NetworkSync} boundary, on a roughly five-minute cadence).
Without intervention, in-flight ACKs dropped during the key swap
trip Pilot's dead-peer detection on otherwise healthy
connections, manifesting at the Entmoot layer as intermittent
gossip failures. The fork's rekey callback resets the keepalive
counter on every \texttt{StateEstablished} connection routed over
the rekeying peer after the in-flight buffer is flushed. The
invariant it restores is that rekey should be transparent to the
reliability state machine; the user-visible stream does not
witness the key change.

The Pilot adapter caches one persistent
\texttt{yamux}~\citep{hashicorpyamux} session per peer and opens
a new yamux stream on that session for every gossip frame. A
first \texttt{Dial} to a given peer pays one Pilot stream dial
plus the yamux handshake; every subsequent \texttt{Dial} returns
a fresh stream with a local \texttt{OpenStream} call. yamux's
built-in keepalive (30-second interval, 10-second write-timeout)
detects silently-wedged sessions and closes them, after which the
next \texttt{Dial} transparently opens a fresh session. A small
\texttt{unblockableConn} shim wraps Pilot's \texttt{driver.Conn}
so that \texttt{Close} propagates a past-time read deadline
ahead of the underlying close: Pilot's driver does not natively
unblock a blocked \texttt{Read} on close, and without this shim
\texttt{yamux.Session.Close} hangs indefinitely waiting for its
\texttt{recvLoop} to exit. This pattern mirrors the
long-lived-framed-stream design used by
\texttt{go-libp2p-pubsub}'s \texttt{comm.go} and by Tailscale
DERP.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  peer/.style={draw, circle, minimum size=8mm, inner sep=0.5pt,
               fill=gray!15},
  stream/.style={-{Latex[length=1.6mm]}, thick},
  tunnel/.style={draw, rounded corners=2pt, line width=0.6mm,
                 gray!55, fill=gray!5},
  clabel/.style={font=\scriptsize, align=center},
]
% Left: naive one-stream-dial-per-frame
\node[peer] (A1) at (0, 0)   {$A$};
\node[peer] (B1) at (1.6, 0) {$B$};
\draw[stream] ($(A1.east) + (0, 0.3)$)  -- ($(B1.west) + (0, 0.3)$);
\draw[stream] ($(A1.east) + (0, 0.1)$)  -- ($(B1.west) + (0, 0.1)$);
\draw[stream] ($(A1.east) + (0, -0.1)$) -- ($(B1.west) + (0, -0.1)$);
\draw[stream] ($(A1.east) + (0, -0.3)$) -- ($(B1.west) + (0, -0.3)$);
\node[clabel] at (0.8, -1.3) {naive\\one stream dial\\per frame};

% Right: one yamux session
\node[peer] (A2) at (3.4, 0) {$A$};
\node[peer] (B2) at (5.6, 0) {$B$};
% Tunnel envelope, taller so three labelled streams fit inside
\draw[tunnel] ($(A2.east)+(0.1, 0.8)$) rectangle ($(B2.west)+(-0.1,-0.8)$);
\draw[stream] ($(A2.east)+(0.15, 0.5)$)
         -- node[above=-0.3mm, font=\tiny, pos=0.5] {\texttt{gossip}}
         ($(B2.west)+(-0.15, 0.5)$);
\draw[stream] ($(A2.east)+(0.15, 0)$)
         -- node[above=-0.3mm, font=\tiny, pos=0.5] {\texttt{fetch\_req}}
         ($(B2.west)+(-0.15, 0)$);
\draw[stream] ($(A2.east)+(0.15,-0.5)$)
         -- node[above=-0.3mm, font=\tiny, pos=0.5] {\texttt{range\_req}}
         ($(B2.west)+(-0.15,-0.5)$);
\node[clabel] at (4.5, -1.3) {Entmoot\\one yamux session,\\many streams};

\end{tikzpicture}
\caption{Stream multiplexing in the Pilot adapter. A naive
adapter would dial a fresh Pilot stream per gossip frame,
paying a SYN handshake on the happy path and the full
$\approx$32-second dial budget on silent loss. Entmoot instead
carries every gossip stream over a single persistent
\texttt{yamux}~\citep{hashicorpyamux} session opened on top of
the Pilot tunnel; subsequent \texttt{gossip}, \texttt{fetch\_req},
and \texttt{range\_req} frames open fresh yamux streams in zero
round trips. yamux's built-in keepalive (30-second interval,
10-second write-timeout) closes silently-wedged sessions so the
next \texttt{Dial} opens a fresh one.}
\label{fig:yamux}
\end{figure}

Dependencies outside the Pilot module are deliberately minimal:
Go's standard library, \texttt{golang.org/x/time/rate} for the
rate-limiter's token-bucket primitive, \texttt{golang.org/x/%
sync/errgroup} for the bounded parallel fanout
(\S\ref{sec:gossip}), and
\texttt{github.com/hashicorp/yamux}~\citep{hashicorpyamux} for
per-peer session multiplexing.

% === 5. Evaluation ===
\section{Evaluation}
\label{sec:evaluation}

We evaluate Entmoot with an end-to-end canary test executed in
three variants (Figure~\ref{fig:canary}). The canary flow is
identical across all three: three gossipers, call them $A$, $B$,
and $C$, initialize
with a shared group whose founder is $A$. $A$ signs \texttt{add}
entries for $B$ and $C$ and replicates the roster to them via the
bootstrap flow. Three messages are then published (first from $A$,
then from $B$, then from $A$ again), each with a distinct topic. We
poll every 100\,ms for convergence, defined as all three members
reporting the same Merkle root and the same message count. A
30-second timeout bounds each variant.

The first variant, \texttt{TestCanaryInMemory}, wires gossipers
together with an in-memory transport implemented as pairs of
\texttt{net.Pipe} connections. It executes in roughly 1.5 seconds of
wall clock and exercises the full stack except the Pilot network
layer. The second variant, \texttt{TestCanaryPilot}, spawns three
sandboxed Pilot daemons as subprocesses with distinct sockets and
identity directories, establishes pairwise trust between them, and
runs the gossipers (as library code) over real encrypted tunnels. It
completes in approximately 12 seconds of wall clock. The breakdown is
dominated by Pilot daemon startup and handshake setup (around 9
seconds combined); once tunnels are in place, message convergence
takes 1.5\,s. The third variant, \texttt{TestCanaryBinary}, spawns
three real \texttt{entmootd} subprocesses alongside the three Pilot
daemons and drives them entirely through the CLI surface
(\texttt{join}, \texttt{publish}, \texttt{tail}, \texttt{query}). It
completes in approximately 14 seconds of wall clock; it is the only
variant that exercises the control-socket IPC path and the
five-command CLI end-to-end, whereas the other two exercise only the
library layer.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  agent/.style={draw, circle, minimum size=10mm, inner sep=0.5pt, fill=gray!10},
  push/.style={-{Latex[length=2mm]}, thick},
  fetch/.style={-{Latex[length=1.8mm]}, dashed},
]
\node[agent] (a) at (2.75, 2.0) {$A$};
\node[agent] (b) at (0, 0)      {$B$};
\node[agent] (c) at (5.5, 0)    {$C$};

% Push (solid): A -> B and A -> C.
\draw[push] (a) -- (b);
\draw[push] (a) -- (c);
% Push labels placed explicitly, outside each arrow.
\node[font=\tiny] at ($(a)!0.5!(b) + (-0.55,0)$) {push id};
\node[font=\tiny] at ($(a)!0.5!(c) + (0.55,0)$)  {push id};

% Fetch (dashed): B -> A and C -> A, curved outward above.
\draw[fetch] (b.north) to[bend left=55]  (a.west);
\draw[fetch] (c.north) to[bend right=55] (a.east);
% Fetch labels placed explicitly, well outside the curved arrows.
\node[font=\tiny] at (-0.15, 1.85) {fetch\_req};
\node[font=\tiny] at (5.65, 1.85)  {fetch\_req};
\end{tikzpicture}
\caption{Canary topology. Founder $A$ publishes a message and
pushes its id to its random fanout (solid). Peers lacking the
body pull it back via \texttt{fetch\_req} (dashed) when the body
is not inlined in the push frame. For three peers with
fanout~$=2$, a single gossip round suffices to reach global
Merkle-root agreement. Small messages (canonical encoding at
most 4\,KiB) inline the body in the push frame and elide the
fetch round-trip; larger messages fall back to the pull path
shown.}
\label{fig:canary}
\end{figure}

Both variants pass reliably across repeated runs. Each publish pushes
its message id to the publisher's two non-self peers; any peer that only
sees the hash pulls the body back via \texttt{fetch\_req}. At fanout two
with three members every publish reaches the entire group in a single
gossip hop, so convergence requires neither anti-entropy nor graft
repair. All three gossipers end on byte-identical Merkle roots.

The three timings above are dominated by Pilot daemon startup and
handshake setup; the fanout-two three-peer convergence ceiling is
bounded below by Pilot's first-contact handshake, not by the
Entmoot dissemination layer. Live operational telemetry on a
three-peer deployment of the asymmetric shape in
Figure~\ref{fig:partialcanary} (one UAE carrier-grade NAT edge,
one Italian residential peer) measures sub-second steady-state
cross-node propagation. The companion evolution
paper~\citep{entmootevolution} records the incident history that
shaped the fanout, retry, and multiplexing policies of
\S\ref{sec:gossip}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  agent/.style={draw, circle, minimum size=12mm, inner sep=1pt,
                fill=gray!10, align=center},
  direct/.style={-, thick},
  blocked/.style={-, thick, dashed, gray!70!black},
  refanout/.style={-{Latex[length=1.6mm]}, thick, dashed,
                   gray!40!black},
]
\node[agent] (vps)    at (2.75, 2.0) {VPS\\{\scriptsize public}};
\node[agent] (phobos) at (0, 0)      {Phobos\\{\scriptsize NATed}};
\node[agent] (laptop) at (5.5, 0)    {laptop\\{\scriptsize CGN}};

\draw[direct] (phobos) -- (vps);
\draw[direct] (vps) -- (laptop);
\draw[blocked] (phobos) -- node[below, font=\scriptsize]
  {blocked (no direct UDP edge)} (laptop);

\draw[refanout] (phobos) to[bend left=20]
  node[above left, font=\scriptsize] {publish} (vps);
\draw[refanout] (vps) to[bend left=20]
  node[above right, font=\scriptsize] {re-fanout} (laptop);
\end{tikzpicture}
\caption{Partial-connectivity canary topology underlying the live
three-peer measurements. VPS is publicly reachable; Phobos
(Italian residential NAT) and laptop (UAE carrier-grade NAT)
cannot open a direct tunnel to each other. Plumtree's first-seen
forwarding rule (\S\ref{sec:gossip}) lets a Phobos publish reach
the laptop with VPS as the re-fanout hop.}
\label{fig:partialcanary}
\end{figure}

This paper does not present load tests at larger scales or quantitative
measurements of rate-limit behavior under attack; both are tracked as
follow-up work.

% === 6. Related Work ===
\section{Related Work}
\label{sec:related}

\citet{calin2026emergent} is the direct empirical peer of this work.
It establishes that self-organizing agent networks of nontrivial size
exist on Pilot today and that their social structure is analyzable from
metadata alone under strong content encryption. Entmoot instantiates a
primitive (group membership with gossip) that the Calin study
explicitly flags as a gap in Pilot's existing surface and a prerequisite
for the cross-network analyses it proposes as future work.

Secure Scuttlebutt~\citep{tarr2019ssb} is the closest published prior
art: a per-identity append-only signed log disseminated by gossip with
anti-entropy repair. Entmoot differs in scope (groups of
trust-gated participants rather than a global pub-sub feed) and in
integrating natively with Pilot's authenticated transport, but the
per-identity signed-log primitive and the gossip+anti-entropy split are
directly analogous. IPFS~\citep{benet2014ipfs} shares with Entmoot both
content-addressing and a DAG-structured history, but it is oriented
toward open distribution of immutable blobs and has no notion of group
scope or trust-gated participation. Certificate Transparency~\citep{laurie2013ct}
directly influences our Merkle log layer: we adopt its domain-separation
convention and the discipline of proving inclusion rather than
recomputing the whole log. We reuse MQTT's topic syntax~\citep{mqtt5}
verbatim, but MQTT's semantics are broker-mediated and a poor fit for a
decentralized overlay. We cite Kademlia~\citep{maymounkov2002kademlia}
as a comparison point for structured peer-to-peer overlays; Entmoot
avoids a DHT and relies on Pilot's pairwise trust for peer discovery.

Classical multi-agent systems
literature~\citep{wooldridge2009,shohamleyton2008,dorri2018} provides
foundational models of agent coordination and game-theoretic
equilibria. It does not directly address the regime Calin documents:
heterogeneous, autonomous populations on a shared overlay where
interaction topologies form without a designer. Entmoot is not a
coordination protocol in that sense; it is a communication substrate
that such coordination can later sit on top of.

% === 7. Discussion and Future Work ===
\section{Discussion and Future Work}
\label{sec:future}

Entmoot ships a deliberately narrow subset of the full design.
The items deferred, with their upgrade paths, are: near-peer
sampling informed by subscriber interest overlap, replacing the
current uniform-random peer sample; cooldown and soft-penalty
schemes on rate-limit breach, replacing hard-disconnect;
DHT-assigned message keepers for retention beyond subscriber
interest, extending the current model where each member stores
only what it cares about; Merkle absence proofs for filtered
subscriber views, extending the positive-inclusion proofs the
implementation provides today; end-to-end group encryption with
shared-key rotation on churn, extending the current plaintext
group content (pairwise tunnels remain encrypted by Pilot); and
multi-admin or quorum-based rosters, relaxing the founder-only
admin rule. Eager-push with lazy
\texttt{ihave}/\texttt{graft}/\texttt{prune} tree repair
(Plumtree; \S\ref{sec:gossip}) and anti-entropy reconciliation
(\S\ref{sec:antientropy}) together close the dissemination gap
flagged in earlier drafts: the former ships messages on the happy
path, the latter heals divergence after partition or slow-start.

When Pilot Networks stabilise beyond early access, the membership
interface can be backed by network membership instead of our signed
roster without protocol-level changes. Whether we adopt that backend
will depend on the shape of Pilot's eventual API and on whether
network membership provides the authorization semantics we need --- in
particular, Pilot Networks' registry-mediated membership lacks the
offline-verifiable signed history Entmoot's roster log provides today,
so a pure substitution would require an auxiliary proof-of-membership
mechanism if offline bootstrap is still a goal. The abstraction is
already in place on the Entmoot side.

Separately, Pilot's pluggable transport
layer~(\S\ref{sec:background}) lets Entmoot run over TCP when a
public-IP peer in the group advertises a TCP endpoint --- the
UDP-hostile-network escape hatch that the Calin study's broader
deployments would otherwise stumble on. A deliberate in-production
test attempting to exercise TCP fallback revealed an orthogonal
property of the v1.1.0 stack we had not explicitly designed for:
with an already-established tunnel, a 57-second iptables-induced
block on outbound UDP between two group members produced \emph{no
transport switch and no application-visible loss}. The message
published by the blocked node during the outage was held in the
local \texttt{yamux} send buffer and delivered to the remote peer
three seconds after the block lifted, in a single 899-byte burst
over the restored UDP path. Pilot's reliable in-tunnel stream
protocol retransmits keepalives indefinitely as long as the tunnel's
crypto state looks alive; yamux's 30-second keepalive + 10-second
write-timeout therefore only trips on \emph{new} dials, not on
buffered traffic over an established session. The practical
implication is that v1.1.0 gives a graceful-degradation window of
roughly one minute for UDP micro-outages, below which the
application observes neither loss nor latency penalty beyond the
outage duration. Forcing an actual TCP fallback therefore requires
either a cold-start dial from the blocked peer (e.g., a pilot-daemon
restart during the block) or a block longer than the buffering
window; quantifying both with a dedicated canary variant is a
natural follow-up.

Entmoot also creates the substrate the empirical questions flagged by
\citet{calin2026emergent} require:
longitudinal analysis of group-membership dynamics, homophily on
topic filters, and community structure beyond the Pilot backbone.
A first natural measurement study would be to deploy Entmoot to a
fraction of the existing Pilot population and observe how rosters
grow, how gossip fanout behaves under real loads, and how topic
filters correlate with agents' declared capability tags. The
Agent-Skills-format skill document shipped alongside the installer
is the concrete enabler for the question ``what do agents do with it
once they have it'': it is the interface agents actually install and
call, rather than a library surface waiting for a developer.

% === 8. Conclusion ===
\section{Conclusion}
\label{sec:conclusion}

Pilot agents already form social networks from pairwise primitives;
they lack the group-scale communication primitives that would let them
deliberate, coordinate, and archive shared knowledge. Entmoot closes
that gap with a Layer-2 protocol that requires only what Pilot already
provides: encrypted unicast tunnels and a pairwise trust system. A
reference implementation demonstrates end-to-end convergence of three
peers over real Pilot tunnels in under twelve seconds of wall clock.
The natural next question is not whether the protocol works, but what
agents do with it once they have it.

% === Bibliography ===
\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
